---
title: "Deconvolution glossary"
author: "Konstantin Zaitsev"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- ## Heterogenity of samples -->
<!-- Samples in biological data are often heterogenious. -->
<!-- Blood for example. -->
<!-- More text incoming. -->

## Mixing process: linear model

Suppose that we want to mix some cell types. Let $W$ be an $n \times c$ matrix where $n$ is the number of genes and $c$ is the number of cell types. Matrix $W$ describes how every gene is expressed in concrete cell type. Let $H$ be an $c \times m$ matrix where $m$ is the number of samples we want to mix. Matrix $H$ describes in what proportions (frequencies) we want to mix our cell types, hence the sum of every column in $H$ has to be equal to one.

Mixing process can be modeled using linear model:
$$ X = W \times H $$
where resulting $n \times m$ matrix describes how each gene is expressed in the resuling heterogenious mix.

For example if we wanted to mix brain and liver in three samples: pure brain, 50% / 50% and pure liver, our matrix H would look like
```{r, echo=FALSE, message=FALSE, warning=FALSE}
H = matrix(c(1.0, 0.0, 0.5, 0.5, 0.0, 1.0), nrow=2, ncol=3)
colnames(H) <- c("Sample 1", "Sample 2", "Sample 3")
rownames(H) <- c("Brain", "Liver")
H
```

if we also knew how some genes are expressed in brain and liver, our matrix W would look like
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(clusdec)
data("datasetLiverBrainLung")
W = 2^datasetLiverBrainLung[1:6, c(4, 1)]
colnames(W) <- c("Brain", "Liver")
W
```
then mixing processing would be modeled as simple matrix multiplication:
```{r, message=FALSE, warning=FALSE}
X <- W %*% H
X
```

## Deconvolution

Deconvolution is opposite process. 

Observed gene expression matrix $X$ is given and we want to find such matrices $W$ and $H$ that

$$ X \approx W \times H $$

where $X$, $W$ and $H$ are nonnegative matricies, where sum of every column of $H$ is close to one. There are a lot of methods that allow us to do such factorization, but we want to find biologically meaningful matrices $W$ and $H$.

If we know more information other than $X$ (e.g. we might know matrices $W$ or $H$, or we might know some signature genes for cell types that are in the mix) we will call it **partial deconvolution problem**. There are a lot of methods developed for solving **partial deconvolution problem**: they use different approaches and require different additional information.

If only X is known we will call it **complete deconvolution problem**. And that's what we are going to solve.

## Signature genes

In terms of deconvolution we are going to call gene **signature** if it is higly expressed in only one cell type in given experiment. Knowledge of such genes allows to perform pretty sensitive deconvolution using DSA algorithm. 

## ClusDec

Main idea of ClusDec is to find such genes that may be signatures for cell types in the mix.

## Accuracy

To decide which combination of clusters is better as putative sigantures we have to introduce a way to estimate accuracy of deconvolution. Lets assume we have already found such matrices $W$ and $H$ that 
$$ X \approx W \times H $$

and we want to estimate how good this factorization is. Let $\tilde{X} = W \times H$, so the most natural way is comparison of the matrices $X$ and $\tilde{X}$. Here we introduce the accuracy of deconvolution:

$$ accuracy(X, \tilde{X}) = || log(X) - log(\tilde{X}) ||_F $$
